Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-07-12 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
edit paper and response letter 2

Empirical simulations are carried out to analyze the relationships between $m$, $c$ and $\frac{\ell}{S}$ (lengthscale $\ell$ relative to the half-range $S$ of the input domain). Figure \ref{fig5_relationships} depicts how these three factors interact and affect the accuracy of the HSGP approximation for a GP with squared exponential, Mat\'ern ($\nu$=3/2), and Mat\'ern ($\nu$=5/2) covariance functions and a single input dimension. More precisely, for a given GP model (with a particular covariance function) with {\color{blue}normalized} length-scale $\frac{\ell}{S}$ (From now to later on we will refer to $\frac{\ell}{S}$ by just $\ell$) and given a boundary factor $c$, Figure \ref{fig5_relationships} shows the minimum number of basis functions $m$ required to obtain an accurate approximation in the sense of satisfying eq. (\ref{eq_diff_covs}). We considered an approximation to be a sufficiently accurate when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq. (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$ such that
%
\begin{equation} \label{eq:rel_total_distance}
 \int | k(\tau) - \tilde{k}_m(\tau)|  \,\mathrm{d}\tau  < 0.01 \int k(\tau) \,\mathrm{d}\tau,
\end{equation}
%
where $\tilde{k}_m$ is the approximate covariance function with $m$ basis functions. Alternatively, these figures can be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be approximated with high accuracy given $m$ and $c$. We obtain the following main conclusions:
%
\begin{itemize}
\item As $\ell$ increases, $m$ required for an accurate approximation decreases. ({\color{blue}Notice that the larger $\ell$ the smoother the functions generated from a covariance function, and vice versa; see covariance functions equations in Section \ref{ch4_sec_cov}.})
\item The lower $c$, the smaller $m$ can and $\ell$ must be to achieve an accurate approximation.
\item For a given $\ell$ there exist a minimum $c$ under which an accurate approximation is never going to be achieved regardless of $m$. This fact can be seen in Figure~\ref{fig5_relationships} as the contour lines which represent $c$ have an end in function of $\ell$ (Valid $c$ are restricted in function of $\ell$). As $\ell$ increases, the minimum valid $c$ also increases.
\end{itemize}

@@

\item[A1.] Set the iteration index to $i=1$. Make an initial guess on the length-scale $\ell^{(i)}$. If there is no useful information available, we recommend to start with a large length-scale, about $\ell^{(i)} \in [0.5, 1]$. ({\color{blue}Remember that $\ell$ refers to the normalized length-scale $\frac{\ell}{S}$.})

\item[A2.] Obtain the minimum valid boundary factor $c^{(i)}$ determined by $\ell = \ell^{(i)}$ for the given kernel as per Section~\ref{sec_num_equations}.

\item[A3.] Obtain the mimimum valid number of basis functions $m^{(i)}$ determined by $\ell = \ell^{(i)}$ and $c = c^{(i)}$ for the given kernel as per Section~\ref{sec_num_equations}.

\item[A4.] Fit an HSGP model using $m^{(i)}$ and $c^{(i)}$ and ensure convergence of the MCMC chains.

\item[A5.] Perform the length-scale diagnostic by checking if $\hat{\ell}^{(i)} - 0.01 \geq \ell^{(i)}$.

If the diagnostic is FALSE, the HSGP approximation is not yet sufficiently accurate. Set $\ell^{(i+1)} = \hat{\ell}^{(i)}$, increase the iteration index $i = i + 1$, and go back to A2.

@@

\item[B2.] Set $m^{(i+1)} = m^{(i)} + 5$ and increase the iteration index $i = i + 1$.

\item[B3.] Obtain the minimum valid boundary factor $c^{(i)}$ determined by $\ell = \hat{\ell}^{(i-1)}$ for the given kernel as per Section \ref{sec_num_equations}.

\item[B4.] Obtain the minimum valid length-scale $\ell^{(i)}$ that can be accurately estimated by $m = m^{(i)}$ and $c = c^{(i)}$ for the given kernel as per Section~\ref{sec_num_equations}.

\item[B5.] Fit an HSGP model using $m^{(i)}$ and $c^{(i)}$ and ensure convergence of the MCMC chains.

\item[B6.] Perform the length-scale diagnostic by checking if $\hat{\ell}^{(i)} - 0.01 \geq \ell^{(i)}$.

Check the stability of both $\hat{\ell}^{(i)}$ and the measures of predictive accuracy relative to the previous iteration.

@@

({\color{blue}In Section \ref{sec_requirements_diagnosis} readers can find a brief remark on the iterations requirements in the diagnosis procedure.})

@@

\subsubsection{Remark on the iterations requirements in the diagnosis procedure}\label{sec_requirements_diagnosis}

{\color{blue}The limitation of the approach that can cause too long computation time does not come from the iterative diagnosis procedure, but from the computational cost of the HMC sampling method to fit the HSGP model at each diagnosis iteration. As said extensively throughout the paper, the computational cost of a single model fit depends ultimately on the non-linearity (wigglyness) of the function to be estimated and the number of dimensions, which affects the HMC sampling of a model fit, but not the number of iterations required to complete the diagnosis procedure. Then, the user has to care about the computational cost of fitting a model, but not about the number of iterations needed to complete the diagnosis procedure, which are few in any case. Few iterations in the diagnosis procedure are sufficient to obtain an accurate model approximation even in multi(low)-dimensional cases (that need a large number of basis functions), as can be seen in Table 1 and Tables 2-4 in the following sections.}

>>>>

Empirical simulations are carried out to analyze the relationships between $m$, $c$ and $\frac{\ell}{S}$ (lengthscale $\ell$ relative to the half-range $S$ of the input domain). Figure \ref{fig5_relationships} depicts how these three factors interact and affect the accuracy of the HSGP approximation for a GP with squared exponential, Mat\'ern ($\nu$=3/2), and Mat\'ern ($\nu$=5/2) covariance functions and a single input dimension. More precisely, for a given GP model (with a particular covariance function) with {\color{blue}normalized} length-scale $\frac{\ell}{S}$ (from now to later on we will refer to $\frac{\ell}{S}$ by just $\ell$) and given a boundary factor $c$, Figure \ref{fig5_relationships} shows the minimum number of basis functions $m$ required to obtain an accurate approximation in the sense of satisfying eq. (\ref{eq_diff_covs}). We considered an approximation to be a sufficiently accurate when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq. (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$ such that
%
\begin{equation} \label{eq:rel_total_distance}
 \int | k(\tau) - \tilde{k}_m(\tau)|  \,\mathrm{d}\tau  < 0.01 \int k(\tau) \,\mathrm{d}\tau,
\end{equation}
%
where $\tilde{k}_m$ is the approximate covariance function with $m$ basis functions. Alternatively, these figures can be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be approximated with high accuracy given $m$ and $c$. We obtain the following main conclusions:
%
\begin{itemize}
\item As $\ell$ increases, $m$ required for an accurate approximation decreases. ({\color{blue}Notice that the larger $\ell$ the smoother the functions generated from a covariance function, and vice versa; see covariance functions equations in Section \ref{ch4_sec_cov}.})
\item The lower $c$, the smaller $m$ can and $\ell$ must be to achieve an accurate approximation.
\item For a given $\ell$ there exist a minimum $c$ under which an accurate approximation is never going to be achieved regardless of $m$. This fact can be seen in Figure~\ref{fig5_relationships} as the contour lines which represent $c$ have an end in function of $\ell$ (Valid $c$ are restricted in function of $\ell$). As $\ell$ increases, the minimum valid $c$ also increases.
\end{itemize}

@@

\item[A1.] Set the iteration index to $k=1$. Make an initial guess on the length-scale $\ell^{(k)}$. If there is no useful information available, we recommend to start with a large length-scale, about $\ell^{(k)} \in [0.5, 1]$. ({\color{blue}Remember that $\ell$ here refers to the normalized length-scale $\frac{\ell}{S}$.})

\item[A2.] Obtain the minimum valid boundary factor $c^{(k)}$ determined by $\ell = \ell^{(k)}$ for the given kernel as per Section~\ref{sec_num_equations}.

\item[A3.] Obtain the mimimum valid number of basis functions $m^{(k)}$ determined by $\ell = \ell^{(k)}$ and $c = c^{(k)}$ for the given kernel as per Section~\ref{sec_num_equations}.

\item[A4.] Fit an HSGP model using $m^{(k)}$ and $c^{(k)}$ and ensure convergence of the MCMC chains.

\item[A5.] Perform the length-scale diagnostic by checking if $\hat{\ell}^{(k)} - 0.01 \geq \ell^{(k)}$.

If the diagnostic is FALSE, the HSGP approximation is not yet sufficiently accurate. Set $\ell^{(k+1)} = \hat{\ell}^{(k)}$, increase the iteration index $k = k + 1$, and go back to A2.

@@

\item[B2.] Set $m^{(k+1)} = m^{(k)} + 5$ and increase the iteration index $k = k + 1$.

\item[B3.] Obtain the minimum valid boundary factor $c^{(k)}$ determined by $\ell = \hat{\ell}^{(k-1)}$ for the given kernel as per Section \ref{sec_num_equations}.

\item[B4.] Obtain the minimum valid length-scale $\ell^{(k)}$ that can be accurately estimated by $m = m^{(k)}$ and $c = c^{(k)}$ for the given kernel as per Section~\ref{sec_num_equations}.

\item[B5.] Fit an HSGP model using $m^{(k)}$ and $c^{(k)}$ and ensure convergence of the MCMC chains.

\item[B6.] Perform the length-scale diagnostic by checking if $\hat{\ell}^{(k)} - 0.01 \geq \ell^{(k)}$.

Check the stability of both $\hat{\ell}^{(k)}$ and the measures of predictive accuracy relative to the previous iteration.

@@

{\color{blue} We have seen this procedure do converge quickly after only few iterations in all the investigated cases (see Section~\ref{sec_cases}). That said, we cannot rule out that there may be cases where convergence is much slower (in terms of number of required diagnostic iterations) to a degree where HSGPs become impractical and an early stopping of the procedure would be advisable.
Importantly though, this scenario should not be confused with the scaling of computational cost due to higher dimensional input spaces, an issue discussed in detail in Section~\ref{sec_cost_multi}. In a nutshell, according to our experiments, increasing the dimensionality of the input space does not increase the number of required diagnostic iterations to a relevant degree but only the computational cost per iteration.}

@@

### 2022-07-12 Paper/Manuscript_Stats&Comp/reply_letter_2nd_STCO.tex
edit paper and response letter 2

\item We have updated the appearance of Figures 1, 2, 3, 4 and 5 (old numbering) in order to make them more compact and thus the overall appearance of the paper more pleasing. The new numbering corresponding to those figures is Fig. 1, 2, 3 and 4.

@@

We are grateful for your valuable comments and suggestions, which have been very helpful to improve the paper, and more generally for your time you have dedicated to reviewing it.

@@

We would also like to thank you for the additional feedback on some of our responses.

@@

Figure 6 may look a bit fuzzy, but it is actually a visual effect because the curves were constructed for discrete values of the number of basis functions $m$ (Y-axis), which makes them look like stair-shaped curves when plotted, and its appearance can be confused with a blurred image.

We have increased the size of the figures in Table 1.

The authors think that the rest of the figures look good in their current form.

@@

The limitation of the approach that can cause too long computation time does not come from the iterative diagnosis procedure, but from the computational cost of the HMC sampling method to fit the HSGP model at each diagnosis iteration, that is, the more input dimensions ($D$) the more multivariate basis functions ($m^*=nm^D$, with $n$=number of points and $m$=number of univariate basis functions), which affects the HMC sampling of a model fit, but not the number of iterations required to complete the diagnosis procedure.

In fact, the diagnosis procedure converges quickly (with few iterations) to an accurate approximation even in multi(low)-dimensional cases (that need a large number of basis functions), as can be seen in Tables 1-5. It is true that more input dimensions may lead to the need for a few more iterations in the diagnosis procedure due to the fact that the diagnostic check is performed individually on each dimension and an accurate approximation of the model will be reached when the diagnostic check passes for all dimensions together. But this fact should only lead to the need for only a couple of additional iterations, since the approach is only suitable for dimensionalities $<= 5$ (Remember that, as stated throughout the paper, the HSGP approach is only suitable for $D<=3$, with (very) wiggly functions, and for $D<=5$, with (very) smooth functions). For larger dimensionalities, the approach can still be used in an additive modeling framework.

As an example, Table 5 summarizes the iterative diagnosis for a 3D function formed by very/moderate wiggly functions in each of its dimensions. For the 4D and 5D cases we do not expect a significant increase in the number of diagnosis iterations, even more so since for the 4D and 5D cases the HSGP model can only fit smooth (or very smooth in 5D) functions.

Therefore, the user should be concerned about the computational cost of fitting a model in a single iteration of the diagnosis procedure, but not about the number of iterations required to complete the diagnosis procedure, which should be few in any case. The computational cost of fitting a single model ultimately depends on the wigglyness of the function and the number of dimensions. We have provided extensive information throughout the paper on the computational requirements for model fitting as a function of wigglyness and number of dimensions, information that is also summarized in Figure 15.

Although that has been said extensively throughout the paper, we have added a remark in Section 4.8.1 and a note at the end of Section 4.5.1, as suggested by the reviewer, which we also think can help to make it more clear.

@@

We are grateful for your valuable comments and suggestions, which have been very helpful to improve the paper, and more generally for your time you have dedicated to reviewing it.

@@

We included the solution of a spline model (a model related to the GP/HSGP model) only in case study I, with no other intention than to show, beside the GP and HSGP solutions, the solution of a spline model, which has similar (of the same order) computational cost and limitations in multidimensional cases. It can be seen that a basic formulation of a spline model causes significantly worse extrapolation accuracy.

@@

As said in the paper, we work with a normalized lengthscale, i.e., a lengthscale $\ell$ relative to the half-range ($S$) of the input domain ($\ell/S$), so the conclusions can be easily referred to any specific scale of the input domain. We have added an additional paragraph to make it more clear in the second-to-last paragraph in Section 4.3.1.

We agree with the reviewer that the $\ell_i$ notation in Section 4.5.1, where $i$ is the subscript for the diagnosis iteration, may conflict with the $\ell_i$ notation used in the article where $i$ is the subscript for dimensions. Therefore, we have changed the notation in Section 4.5.1 from $\ell_i$ to $\ell^{(i)}$ to differentiate it from the notation in the rest of the article where subscripts refer to dimensions.

>>>>

\item We have updated the appearance of Figures 1, 2, 3, 4, and 5 (old numbering) in order to make them more compact and thus the overall appearance of the paper more pleasing. The new numbering corresponding to those figures is now Fig. 1, 2, 3, and 4.

@@

Thank you again for your valuable comments and suggestions.

%We are grateful for your valuable comments and suggestions, which have been very helpful to improve the paper, and more generally for your time you have dedicated to reviewing it. We would also like to thank you for the additional feedback on some of our responses.

@@

@@

Figure 6 may look a bit fuzzy, but it is actually a visual effect because the curves were constructed for discrete values of the number of basis functions $m$ (Y-axis), which makes them look like stair-shaped curves when plotted, and its appearance can be confused with a blurred image. We think the resolution of the figure itself is sufficient. We have increased the size of the figures in Table 1 and have checked and updated the other figures as we saw fit.

%The authors think that the rest of the figures look good in their current form.

@@

Based on the evidence that we have at this point, we can unfortunately not give evidence-based suggestions as to when to stop pre-convergence and try something else because we have seen only very few iterations to be required in all the investigated cases. With regard to high-dimensional cases, the main problem would not be more iterations (as far as we have seen) but the computational cost of each iteration. More concretely, as we discuss in detail in the paper, the more input dimensions ($D$) we have, the more multivariate basis functions we get ($m^*=nm^D$, with $n$=number of points and $m$=number of univariate basis functions), which affects the HMC sampling of a model fit (per iteration), but not (in a relevant manner) the number of iterations required to complete the diagnostic procedure. We have now clarified these points at the end of Section 4.5.1.

%The limitation of the approach that can cause too long computation time does not come from the iterative diagnostic procedure, but from the computational cost of the HMC sampling method to fit the HSGP model at each diagnostic iteration, that is, the more input dimensions ($D$) the more multivariate basis functions ($m^*=nm^D$, with $n$=number of points and $m$=number of univariate basis functions), which affects the HMC sampling of a model fit, but not the number of iterations required to complete the diagnostic procedure.

%In fact, the diagnostic procedure converges quickly (with few iterations) to an accurate approximation even in multi(low)-dimensional cases (that need a large number of basis functions), as can be seen in Tables 1-5. It is true that more input dimensions may lead to the need for a few more iterations in the diagnostic procedure due to the fact that the diagnostic check is performed individually on each dimension and an accurate approximation of the model will be reached when the diagnostic check passes for all dimensions together. But this fact should only lead to the need for only a couple of additional iterations, since the approach is only suitable for dimensionalities $<= 5$ (Remember that, as stated throughout the paper, the HSGP approach is only suitable for $D<=3$, with (very) wiggly functions, and for $D<=5$, with (very) smooth functions). For larger dimensionalities, the approach can still be used in an additive modeling framework.

%As an example, Table 5 summarizes the iterative diagnostic for a 3D function formed by very/moderate wiggly functions in each of its dimensions. For the 4D and 5D cases we do not expect a significant increase in the number of diagnostic iterations, even more so since for the 4D and 5D cases the HSGP model can only fit smooth (or very smooth in 5D) functions.

%Therefore, the user should be concerned about the computational cost of fitting a model in a single iteration of the diagnostic procedure, but not about the number of iterations required to complete the diagnostic procedure, which should be few in any case. The computational cost of fitting a single model ultimately depends on the wigglyness of the function and the number of dimensions. We have provided extensive information throughout the paper on the computational requirements for model fitting as a function of wigglyness and number of dimensions, information that is also summarized in Figure 15.

%Although that has been said extensively throughout the paper, we have added a remark in Section 4.8.1 and a note at the end of Section 4.5.1, as suggested by the reviewer, which we also think can help to make it more clear.

@@

%We are grateful for your valuable comments and suggestions, which have been very helpful to improve the paper, and more generally for your time you have dedicated to reviewing it.

Thank you again for your valuable comments and suggestions.

@@

We apologize for not making a stronger case in terms of performed (and not performed) comparisons earlier. We are aware that this paper could be extended in various different valuable directions, as is probably the case with most papers. One of such directions is of course to add more comparisons with other methods.

In this paper, we have focussed on the empirical accuracy of the HSGP approximation of exact GPs as you are aware. Including the spline in one of the case studies was to show that excessively applied methods that are similar in some ways (here: reduced-rank thin-plate splines, a method whose corresponding papers have been cited thousands of times already), does not perform better out of the box than our HSGPs. But we understand and agree that including this particular spline here in this particular case study is not a strong case in terms of comparison with other methods. Accordingly, we decided to remove it to not invoke a false impression of our focus or claims.

We are not sure whether we can convince you of the sensibility and sufficiency of the focus of our paper that we have chosen. But what we can say is that based on the informal feedback from the community we have recieved so far, the other reviewers' and also the editor's comments, as well as over 20 citation of the corresponding preprint already, a lot of people see value in our work and its existing focus. There is certainly a lot of value in comparing HSGPs with competing methods other than exact GPs, but we do not see a way to make this part of the current paper, not only in terms of paper space, but also in terms of time, personnel, and monetary resources we currently have available to work on this topic.

%We included the solution of a spline model (a model related to the GP/HSGP model) only in case study I, with no other intention than to show, beside the GP and HSGP solutions, the solution of a spline model, which has similar (of the same order) computational cost and limitations in multidimensional cases. It can be seen that a basic formulation of a spline model causes significantly worse extrapolation accuracy.

@@

We apologize for creating confusion with our notation here. We work with a normalized lengthscale, i.e., a lengthscale $\ell$ relative to the half-range ($S$) of the input domain ($\ell/S$), so the conclusions can be easily referred to any specific scale of the input domain. We have now clarified the notation by using different symbols for lengthscale and normalized lengthscale throughout the paper.

%added an additional paragraph to make it more clear in the second-to-last paragraph in Section 4.3.1.

Thank you for also your comment about the $\ell_i$ notation in Section 4.5.1, where $i$ is the subscript for the diagnostic iteration, conflicting with the $\ell_i$ notation used in the article where $i$ is the subscript for dimensions. Therefore, we have changed the notation in Section 4.5.1 from $\ell_i$ to $\ell^{(k)}$ to clearly differentiate it from the notation in the rest of the article where subscripts refer to dimensions.
